---
title: "Class 9 mini project"
author: 'Natalie Avina (PID: A15590695)'
date: "10/26/2021"
output: html_document
---

```{r}

# Save your input data file into your Project directory
fna.data <- "WisconsinCancer (1).csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv(fna.data, row.names=1)

View(wisc.df)

```

Because we will not be using the first column "diagnosis" we will remove it and make a new data frame:

```{r}
# We can use -1 here to remove the first column
wisc.data <- wisc.df[,-1]
wisc.data
```

We can store that first column here:
```{r}
diagnosis <- wisc.df$diagnosis
```

>Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

>Q2. How many of the observations have a malignant diagnosis?

```{r}
length(grep("M", diagnosis, ignore.case = FALSE, perl = FALSE, value = FALSE,
     fixed = FALSE, useBytes = FALSE, invert = FALSE))
```

```{r}
table(diagnosis)

```

>Q3. How many variables/features in the data are suffixed with _mean

```{r}
features <- colnames(wisc.data)
length(grep("_mean",features))
```

**PRINCIPAL COMPONENT ANALYSIS**

```{r}
colMeans(wisc.data)

apply(wisc.data,2,sd)
```
We have to scale the data with scale=TRUE to make the data comparable with each other due to different units
```{r}
wisc.pr <- prcomp( wisc.data, scale=TRUE )
```

```{r}
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
      44.27%

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?
     3

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?
     7

Creating a plot of our wisc.pr

```{r}
biplot(wisc.pr)
```
>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
It's difficult to understand because there is a lot of data measured in different ways; it overlaps too much for us to read it.

```{r}

# Scatter plot observations by components 1 and 2
#as.factor changes diagnosis into factors
plot( wisc.pr$x[,1:2], col = as.factor(diagnosis) , 
     xlab = "PC1", ylab = "PC2")
```

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?
The data clusters about PC3

```{r}

# Repeat for components 1 and 3
plot(wisc.pr$x[,1], wisc.pr$x[,3], col = as.factor(diagnosis), 
     xlab = "PC1", ylab = "PC3")
```

GGPLOT
```{r}
# Create a data.frame for ggplot
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis

# Load the ggplot2 package
library(ggplot2)


```

```{r}
# Make a scatter plot colored by diagnosis
ggplot(df) + 
  aes(PC1, PC2, col=diagnosis) + 
 geom_point()
```

```{r}
# Calculate variance of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)
```

```{r}
# Variance explained by each principal component: pve
pve <- pr.var / sum(pr.var)

# Plot variance explained for each principal component
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

Alternative Scree Plot

```{r}
# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

```{r}
## ggplot based graph
install.packages("factoextra")
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```



>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

```{r}
wisc.pr$rotation["concave.points_mean",1]
```


>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?
5

```{r}

var <- summary(wisc.pr)
var
```

Heirarchical CLustering

```{r}
# Scale the wisc.data data using the "scale()" function
data.scaled <- scale(wisc.data)

```


```{r}
# Scale the wisc.data data using the "scale()" function
data.dist <- dist(data.scaled)

```


```{r}
wisc.hclust <- hclust(data.dist, method="complete")
plot(wisc.hclust)
abline(h=19, col="red", lty=2)
```

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)

```

```{r}
table(wisc.hclust.clusters, diagnosis)
```

>Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

```{r}
wisc.hclust.clusters2 <- cutree(wisc.hclust, k=3)
table(wisc.hclust.clusters2, diagnosis)
```


>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.
Ward.D2 method minimizes variance between the clusters so it is more likely to group them together by diagnosis

```{r}
wisc.hclust2 <- hclust(data.dist, method="ward.D2")
plot(wisc.hclust2)
abline(h=19, col="red", lty=2)
```
# 4. Kmeans Clustering
```{r}
wisc.km <- kmeans(data.scaled, centers= 2, nstart= 20)
```

```{r}
table(wisc.km$cluster, diagnosis)
```



# 5. *COMBINING METHODS*

Take results of PCA analysis and cluster in this space 'wisc.pr$x'
```{r}
summary(wisc.pr)
```

```{r}
wisc.pc.hclust <- hclust( dist( wisc.pr$x[,1:3] ), method= "ward.D2")
```

Plot dendrogram
```{r}
plot ( wisc.pc.hclust )
abline(h=60, col="red")
```

Cut tree into k=2 groups
```{r}
grps <-  cutree(wisc.pc.hclust, k=2)
table(grps)
```

How do we check if these two groups are dominated by a particular diagnosis?
Cross tanle comare of diagnosis and cluster groups
```{r}
table(diagnosis, grps)
```
```{r}
plot(wisc.pr$x[,1:2], col=grps)
```

```{r}
plot(wisc.pr$x[,1:2], col=as.factor(diagnosis))
```

```{r}
g <- as.factor(grps)
levels(g)
```

```{r}
g <- relevel(g,2)
levels(g)
```

```{r}
plot(wisc.pr$x[,1:2], col=g)
```

```{r}
wisc.pr.hclust <- hclust(dist(wisc.pr$x[, 1:7]), method="ward.D2")
```

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

>Q15. How well does the newly created model with four clusters separate out the two diagnoses?

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```
```{r}
329/ (329+28)
```
```{r}
188/(188+24)
```


>Q16. How well do the k-means and hierarchical clustering models you created in previous sections (i.e. before PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.km$cluster and wisc.hclust.clusters) with the vector containing the actual diagnoses

```{r}
table(wisc.km$cluster, diagnosis)

```
```{r}
table(wisc.hclust.clusters, diagnosis)
```

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```

# 6. Sensitivity vs Specificity
  Kmeans:
```{r}
 #Sensitivity: TP/(TP+FN) refers to true positives of malignant diagnoses contained in the cluster compared to known malignant diagnoses
175/(175+37)

# Specificity: TN/(TN+FN)

343/(343+37)
```
  HClust:
```{r}
#Sensitivity: TP/(TP+FN)

# Specificity: TN/(TN+FN)

```


Pr.Hclust
```{r}
#Sensitivity: TP/(TP+FN)
188/188
# Specificity: TN/(TN+FN)

```


> Q17. Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

```{r}

```

