---
title: "Machine Learning 1"
author: 'Natalie Avina (PID: A15590695)'
date: "10/21/2021"
output:
  pdf_document: default
  html_document: default
---

Clustering Methods

#Kmeans clustering

The function in base R to do Kmeans clustering is called 'kmeans()'

First we make up some data where we know what the answer should be

```{r}
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
x
plot(x)

```

> Q. Can we use kmeans() to cluster this data 
setting k to 2 and nstart to 20

```{r}
km <- kmeans(x, center=2, nstart= 20)
km
```

>Q. How many points are in each cluster?

```{r}
km$size
```

>q. What 'component of your result object details cluster assignment/membership?

```{r}
km$cluster
```


> Q. What component of your result object details cluster center?

```{r}
km$centers
```

> Q. Plot x colored by the kmeans cluster assignment and add cluster centers as blue points.

```{r}
plot(x, col=km$cluster, )
points(km$centers, col="blue", pch=15, cex=2)
```

# Hierarchical Clustering

A limitation with k-means is that we have to specify K (or the number of clusters we want). 

```{r}
#hclust(d, method= , members=) where d is the output of 'dist()'

hc <- hclust(dist(x))
hc
```

The plot method for hclust() is a dendrogram.

```{r}
plot(hc)
```

To get our cluster membership vector we have to "cut" the tree where we want.

```{r}
cutree(hc, h=6)

```

#if you use cutree(hc, k= 'number') you can set k to a number of different groups so that the tree is cut at those groups rather than at a given height

```{r}
grps <- cutree(hc, k=2)
```


Make results plot

```{r}
plot(x, col=grps)
```



**PART 2** 
#Principal Component Analysis

```{r}
url <- "https://tinyurl.com/UK-foods"
y <- read.csv(url, row.names = 1)
```


```{r}
y
```

>Q1. How many rows and columns are in your new data frame named x? What R functions could you use to answer this questions?

```{r}
dim(y)
ncol(y)
nrow(y)
```

We have one extra column that we need to fix so we have to fix that.
```{r}
head(y)
 #views first six rows, tail() views last six rows
```

Be careful because certain commands will remove one column each time you run them such as: 

```{r}
#rownames(x) <- x[,1]
#x <- x[,-1]
#head(x)
```


To fix this we use read.csv(url, row.names=1) instead back up top

>Q2. Which approach to solving the ‘row-names problem’ mentioned above do you prefer and why? Is one approach more robust than another under certain circumstances?
  
  read.csv(url, row.names=1) because it will not remove a row every time you run the code. It is more robust because it can apply to other circumstances and won't eliminate data.
  
  
Barplot

```{r}
barplot(as.matrix(y), beside=T, col=rainbow(nrow(y)))
```

>Q3: Changing what optional argument in the above barplot() function results in the following plot?

```{r}
#Beside=T makes it so that columns are juxtaposed as bars and FALSE makes it so that they are stacked
barplot(as.matrix(y), beside=FALSE, col=rainbow(nrow(y)))
```


>Q5: Generating all pairwise plots may help somewhat. Can you make sense of the following code and resulting figure? What does it mean if a given point lies on the diagonal for a given plot?

```{r}
pairs(y, col=rainbow(10), pch=16)
```

>Q6. What is the main differences between N. Ireland and the other countries of the UK in terms of this data-set?
  
  They consume more potatoes and less fresh fruit than other countries


```{r}
pca <- prcomp( t(y) )
summary(pca)
```


```{r}
attributes(pca)
```


```{r}
plot(pca$x[,1], pca$x[,2])
```

>Q7. Complete the code below to generate a plot of PC1 vs PC2. The second line adds text labels over the data points

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(y))
```

>Q8. Customize your plot so that the colors of the country names match the colors in our UK and Ireland map and table at start of this document

```{r}
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2", xlim=c(-270,500))
text(pca$x[,1], pca$x[,2], colnames(y), col=rainbow(10))
```

```{r}
v <- round( pca$sdev^2/sum(pca$sdev^2) * 100 )
v
```


```{r}
z <- summary(pca)
z$importance
```

```{r}
barplot(v, xlab="Principal Component", ylab="Percent Variation")
```

```{r}
## Lets focus on PC1 as it accounts for > 90% of variance 
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,1], las=2 )
```


>Q9: Generate a similar ‘loadings plot’ for PC2. What two food groups feature prominantely and what does PC2 maninly tell us about?
  The two food groups are Potatoes and Soft Drinks. PC2 mainly tells us about the second highest amount of variation between the countries. 

```{r}
par(mar=c(10, 3, 0.35, 0))
barplot( pca$rotation[,2], las=2 )
```

#Another way to do this is to do a BC plot

```{r}
## The inbuilt biplot() can be useful for small datasets 
biplot(pca)
```

## PCA of RNA-Seq Data

```{r}
url2 <- "https://tinyurl.com/expression-CSV"
rna.data <- read.csv(url2, row.names=1)
head(rna.data)
```

>Q10: How many genes and samples are in this data set?

```{r}
#The number of genes is the number of rows and the number of samples are columns
ncol(rna.data)
nrow(rna.data)
```

There are 100 genes and 10 samples.

```{r}
## Again we have to take the transpose of our data 
pca <- prcomp(t(rna.data), scale=TRUE)
 
## Simple un polished plot of pc1 and pc2
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```


```{r}
summary(pca)
```

A barplot of this data:
```{r}

plot(pca, main="Quick scree plot")
```

```{r}
## Variance captured per PC 
pca.var <- pca$sdev^2

## Percent variance is often more informative to look at 
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)
pca.var.per
```


Beautifying our plot

```{r}

## A vector of colors for wt and ko samples
colvec <- colnames(rna.data)
colvec[grep("wt", colvec)] <- "red"
colvec[grep("ko", colvec)] <- "blue"

plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
     xlab=paste0("PC1 (", pca.var.per[1], "%)"),
     ylab=paste0("PC2 (", pca.var.per[2], "%)"))

text(pca$x[,1], pca$x[,2], labels = colnames(rna.data), pos=c(rep(4,5), rep(2,5)))
```

```{r}
library(ggplot2)

df <- as.data.frame(pca$x)

# Our first basic plot
ggplot(df) + 
  aes(PC1, PC2) + 
  geom_point()
```




```{r}
# Add a 'wt' and 'ko' "condition" column
df$samples <- colnames(rna.data) 
df$condition <- substr(colnames(rna.data),1,2)

p <- ggplot(df) + 
        aes(PC1, PC2, label=samples, col=condition) + 
        geom_label(show.legend = FALSE)
p
```

#Finalized plot with new fixes

```{r}
p + labs(title="PCA of RNASeq Data",
       subtitle = "PC1 clealy seperates wild-type from knock-out samples",
       x=paste0("PC1 (", pca.var.per[1], "%)"),
       y=paste0("PC2 (", pca.var.per[2], "%)"),
       caption="BIMM143 example data") +
     theme_bw()
```



