---
title: "Class 17 Mini Project"
author: 'Natalie Avina (PID: A15590695)'
date: "11/23/2021"
output: html_document
---

```{r}
# Import vaccination data
vax <- read.csv( "covid19vaccinesbyzipcode_test.csv" )
head(vax)
```

```{r}
library(skimr)
```

```{r}
skimr::skim(vax)
```


## Ensure the date colume is useful 
We will use the **lubridate** package 
```{r}
#install.packages("lubridate")
library(lubridate)
```
```{r}
today()
```

Let's convert the data in the 'as_of_date' column to lubridate format

```{r}
vax$as_of_date <- ymd(vax$as_of_date)
```

```{r}
today() - vax$as_of_date[1]
```

```{r}
nrow(vax)
```

> When was the last day of entry?

```{r}
today()- vax$as_of_date[nrow(vax)]
```

>Q9. How many days b/t the first and last entry in the data set?

To give us the span of days that the data is from:
```{r}
vax$as_of_date[nrow(vax)] - vax$as_of_date[1]
```

> Q10. How many unique dates are in the data set?

```{r}
length(unique(vax$as_of_date))
```
This makes sense
```{r}
46*7
```

```{r}
install.packages("zipcodeR")
library(zipcodeR)
```

```{r}
geocode_zip('92037')
```

```{r}
zip_distance('92037','92109')
```

```{r}
reverse_zipcode(c('92037', "92109") )

```

```{r}
#zipdata <- reverse_zipcode( vax$zip_code_tabulation_area )
```



Let's focus on San Diego County:

```{r}
library(dplyr)

sd <- filter(vax, county == "San Diego")

nrow(sd)
```

```{r}
sd.10 <- filter(vax, county == "San Diego" &
                age5_plus_population > 10000)
```

>Q11. How many distinct zip codes are listed for San Diego County? 

```{r}
length(unique(sd$zip_code_tabulation_area))
```


>Q12. What San Diego County Zip code area has the largest 12 + Population in this dataset? 

```{r}
ind <- which.max(sd$age12_plus_population)
sd[ind,]
```

> What is the population in the 92037 zipcode area?

```{r}
filter(sd, zip_code_tabulation_area == "92037")[1,]
```

>Q13. what is the overall avg "Perent of Population Fully Vaccinated" value for all san diego "County" as of "2021-11-09"?

```{r}
sd.now <- filter(sd,as_of_date== "2021-11-09")
mean(sd.now$percent_of_population_fully_vaccinated, na.rm= TRUE)
```

We can look at the 6-number summary
```{r}
summary( sd.now$percent_of_population_fully_vaccinated )
```

```{r}
library(ggplot2)
```

```{r}
ggplot(sd.now, aes(percent_of_population_fully_vaccinated))+
  geom_histogram(bins=15)
```

What about 92037 -UCSD/La Jolla?
```{r}
filter(sd.now, zip_code_tabulation_area == "92037")

```
```{r}
ucsd <- filter(sd, zip_code_tabulation_area=="92037")
ucsd[1,]$age5_plus_population
```


```{r}
ggplot(ucsd) +
  aes(x= as_of_date,
      y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated")
```

##Comparing 92037 to other similarly sized areas

```{r}
vax.36 <- filter(vax, age5_plus_population > 36144 &
                as_of_date == "2021-11-16")
head(vax.36)
```

>Q16. Calculate the mean “Percent of Population Fully Vaccinated” for ZIP code areas with a population as large as 92037 (La Jolla) as_of_date “2021-11-16”. Add this as a straight horizontal line to your plot from above with the geom_hline() function?

```{r}
mean()
```

```{r}
ggplot(ucsd) +
  aes(x= as_of_date,
      y = percent_of_population_fully_vaccinated) +
  geom_point() +
  geom_line(group=1) +
  ylim(c(0,1)) +
  labs(x = "Date", y="Percent Vaccinated") +
  geom_hline(yintercept = 0.66)
```

```{r}
ucsd[1,]$age5_plus_population
```

First we need to subset 'vax' dataset to include only zipcode areas with a population as large as 92037
```{r}
vax.36.all <- filter(vax, age5_plus_population > 36144)
nrow(vax.36.all)
```

```{r}
length(unique(vax.36.all$zip_code_tabulation_area))
```


>Q20.

Let's make a final figure that shows all these ZIP areas
```{r}
ggplot(vax.36.all) + 
  aes(x= as_of_date,
      y = percent_of_population_fully_vaccinated, group=zip_code_tabulation_area)+
  geom_line(alpha=0.2) + 
  geom_hline(yintercept = 0.66, col="red")
```

>Q21. How do you feel about traveling for Thanksgiving and meeting for in-person class next Week?


